The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-23 conf_article/neon.tex
Defined Fanions

\begin{abstract}
In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the Ky Fan k-norms, their affine combinations with the Frobenius norms, and corresponding duals to develop a new family of Muon-like algorithms. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings.
\end{abstract}

@@

    {\bf $\normf{\mM^{k}}$ and NSGD}

@@

    {\bf $\norms{\mM^{k}}$ and Muon}
    \begin{lemma}\label{lemma:muon_update}
        When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

@@

    {\bf $\normn{\mM^{k}}$ and Neon}
    \begin{lemma}\label{lemma:neon_update}
        When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

@@

    {\bf $\normfstar{\mM^{k}}^\dagger$ and F-Muon}
        We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

        where $\alpha \in [0, 1]$ defines a specific norm of $F*$-family.

@@

    {\bf $\normftwo{\mM^{k}}^\dagger$ and F-Neon}
        We define $\normftwo{\cdot}$ as a convex combination of $\norms{\cdot}$ and $\normf{\cdot}$:

@@

        where $\alpha \in [0, 1]$ defines a specific norm of F2-family.

@@

    {\bf $\normkfk{\mM^k}^\dagger$ and Muon, Neon, and Centralized Dion without error feedback}
        We remind the reader that the Ky Fan k-norm (\cite{bhatia2013matrix}, p. 92), which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proved that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \cite{bhatia2013matrix}, p. 96). Special cases of the Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

            \begin{equation}\label{eq:ky_fan_update}
                \mX^{k+1} = \mX^{k} - \eta \sum_{i=1}^{k}u_i v_i^\top
            \end{equation}
        \end{lemma}

        This lemma recovers the updates of Muon, when $k=\min\{m, n\}$, of Neon, when $k=1$, and of Dion \cite{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback.

        Moreover, one can consider F-KF-k-norm = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback.

@@

Here we provide the derivation of these norms and plot them in the case of $2\times2$ matrices.

\subsection{Visualization of different matrix norms}

TO-DO: Vibe-code F2, F* norm plots and their duals.

1-balls in $l_\infty$, $l_1$ and $l_2$ norms are well-known from textbooks. But what about Ky Fan $k$-norm? How can it be represented?

>>>>

\begin{abstract}
In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the Ky Fan k-norms, their affine combinations with the Frobenius norms, and corresponding duals to develop a new family of Muon-like algorithms we name {\it Fanions}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings.
\end{abstract}

@@

    \subsection{\texorpdfstring{$\normf{\mM^{k}}$ and Normalized SGD}{NSGD}}

@@

    \subsection{\texorpdfstring{$\norms{\mM^{k}}$ and Muon}{Muon}}
    \begin{lemma}\label{lemma:muon_update}
        When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

@@

    \subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
    \begin{lemma}\label{lemma:neon_update}
        When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

@@

    \subsection{\texorpdfstring{$\normfstar{\mM^{k}}^\dagger$ and F-Muon}{F-Muon}}
        We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

        where $\alpha \in [0, 1]$ defines a specific norm from the $F*$-family.

@@

    \subsection{\texorpdfstring{$\normftwo{\mM^{k}}^\dagger$ and F-Neon}{F-Neon}}
        We define $\normftwo{\cdot}$ as a convex combination of $\norms{\cdot}$ and $\normf{\cdot}$:

@@

        where $\alpha \in [0, 1]$ defines a specific norm from the F2-family.

@@

    \subsection{\texorpdfstring{$\normkfk{\mM^k}^\dagger$ and Muon, Neon, and Centralized Dion without error feedback}{Muon, Neon, and Centralized Dion without error feedback}}
        We remind the reader that the Ky Fan k-norm (\cite{bhatia2013matrix}, p. 92), which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proved that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \cite{bhatia2013matrix}, p. 96). Special cases of the Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

            \begin{equation}\label{eq:ky_fan_update}
                \mX^{k+1} = \mX^{k} - \eta \sum_{i=1}^{k}u_i v_i^\top
            \end{equation}
        \end{lemma}
        In the section {\it The Matrix side of updates}, we will discuss how to compute the updates efficiently.

        We introduce the family of {\it Fanions}, which consists of {\it $k$-Fanions}, lmo-based algorithms under the $\normkfk{\mM^k}^\dagger$ norms. By this terminology and the lemma, Muon is a $\min\{n,m\}$-Fanion, while Neon is an 1-Fanion. Moreover, the centralized version of rank-$r$ Dion \cite{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a $r$-Fanion.

        In addition, one can consider F-KF-k-norm = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback. The resulting family can be named {\it F-Fanions}.

@@

\subsection{Visualization of different matrix norms}
\subsubsection{Duals to F* and F2 norms}
In \ref{fig:fduals} we plot these norms. On x-axis and y-axis, there are singular values $\sigma_1$, $\sigma_2$ respectively of a matrix from $\Rmn$ with $\min\{m,n\}=2$.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fstardualball.pdf}
        \caption{The dual to F* norm for different $\alpha$}
        \label{fig:fstardual}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{ftwodualball.pdf}
        \caption{The dual to F2 norm for different $\alpha$}
        \label{fig:ftwodual}
    \end{subfigure}
    \caption{The duals to F* and F2 norms for different $\alpha$}
    \label{fig:fduals}
\end{figure}

\subsubsection{The Ky Fan norm and its dual}
1-balls in $l_\infty$, $l_1$ and $l_2$ norms are well-known from textbooks. But what about the Ky Fan $k$-norm? How can it be represented?
